---
title: "Coursera - Practical Machine Learning"
author: "Dominik Koch"
date: "Sunday, June 21, 2015"
output: html_document
---

-----

### Story

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

### Data

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

### Task

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 

1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders :-).
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details. 

Due to security concerns with the exchange of R code, your code will not be run during the evaluation by your classmates. Please be sure that if they download the repo, they will be able to view the compiled HTML version of your analysis. 

-----

### General preparation

At the beginning of the code all required R Packages are loaded. The working
directory is needed for creating the data folder. And the seed is set for 
reproducibility.

```{r, message = FALSE}
# Import packages
library(knitr)
library(caret)

# Working directory
wd <- getwd()

# Reproducibility
set.seed(42)
```

### Download data

Download the csv files in the data folder. If there is no *data* folder yet, it
will be created. The data will only be downloaded once.

```{r}
# Create folder
if(!file.exists(file.path(wd,"data"))){
  dir.create(file.path(wd,"data"))
}

# Download csv files
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
if(!file.exists(file.path(wd,"data","training.csv"))){ 
  download.file(url, file.path(wd,"data","training.csv")) 
}

url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
if(!file.exists(file.path(wd,"data","testing.csv"))){ 
  download.file(url, file.path(wd,"data","testing.csv")) 
}

# Check data
list.files(file.path(wd,"data"))
```

### Import data 

In the next step the csv files will be imported

```{r}
data.training <- read.table("data\\training.csv", sep=",", header = TRUE)
data.coursera <- read.table("data\\testing.csv",  sep=",", header = TRUE)

dim(data.training)
dim(data.coursera)
#colnames(data.training)
```

The training data set consists of `r nrow(data.training)` observations and `r ncol(data.training)` 
variables. The test data set contains only `r nrow(data.coursera)` observations and `r ncol(data.coursera)` variables. We want to predict the variable called **classe**. In the data.coursera set the outcome was replaced by the variable **problem_id**.

### Clean data 

In this section all useless variables will be filtered out. At first some general information
like timestamps or simple row numbers because these variables won't contribute to the classification.

```{r, echo = FALSE}
for(i in 1:6) print(colnames(data.training)[c(1,3,4,5,6,7)][i])
```

```{r}
data.training <- data.training[,-c(1,3,4,5,6,7)]
```

Next, I want to eliminate variables which contain missing values. Out of the remaining `r ncol(data.training)` variables, `r sum(colSums(is.na(data.training)) == 0)` contain missing values. It is allowed drop these features because imputation is not recommended (19216 values missing out of 19622).

```{r}
data.training <- data.training[, colSums(is.na(data.training)) == 0] 
```

In the next step I only want to drop near zero variance predictors. These predictors only have one unique value or have both of the following characteristics:

- they have very few unique values relative to the number of samples
- the ratio of the frequency of the most common value is large

This is true for:
```{r}
nzv <- nearZeroVar(data.training, saveMetrics = TRUE)
nzv[nzv$nzv == TRUE, ]
```

```{r}
nzv <- nearZeroVar(data.training)
data.training <- data.training[, -nzv]
```

After dropping near zero variance predictors there are `r nrow(data.training)` observations and `r ncol(data.training)` variables left.


### Split data 

Now it is time to split the available data in a training (70%) and a testing (30%) partition. The training set will be used to train the model an the testing set is for validating the model on new data. It is very important to use new data for the validation in order to recieve unbiased performance results.

```{r}
inTrain <- createDataPartition(y=data.training$classe, p=0.7, list=FALSE)
data.testing <-  data.training[-inTrain,]
data.training <- data.training[inTrain,]
```

### Build Model 

For predicting the different activities I used a **Random Forest Algorithm**. The great benefit of the random forest is the integrated variable selection and its high accuracy. The difficult interpretation is bearable because I'm not interested in the specific effect of the different variables.
In order to prevent the model from overfitting I used an **5 fold Cross Validation** in the model tuning process. For runtime issues I set the **number of trees** to 50. With higher numbers you can improve the prediction (for example ntree = 250).

```{r}
fitControl <- trainControl(
  method = "cv",
  number = 5)
  
fitMod <- train(classe ~ ., 
                method = "rf", 
                trainControl = fitControl, 
                data = data.training, 
                ntree = 50)

fitMod
```

### Evaluate Model

After building the model I want to know its performance. Therefore I use the testing data set I put aside earlier.

```{r}
evalMod <- predict(fitMod, newdata = data.testing)
confusionMatrix(data.testing$classe, evalMod)
```

The confusion Matrix seems pretty good. Only few observations were predicted into the wrong class. Also the accuracy with `r postResample(evalMod, data.testing$classe)[1]` is pretty high.

```{r}
accuracy <- postResample(evalMod, data.testing$classe)
accuracy
```

The expected out of sample error is calculated via 1 minus the accuracy.

```{r}
eoo.sampleerror <- 1 - as.numeric(accuracy[1])
eoo.sampleerror
```



# Prediction of the 20 test cases
-----

Last but not least I still have to predict the 20 test cases. 

```{r}
pred.coursera <- predict(fitMod, newdata = data.coursera)
pred.coursera
```

And write my predictions to txt-files.

```{r}
if(!file.exists(file.path(wd,"submission"))){
  dir.create(file.path(wd,"submission"))
}

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("submission\\problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(pred.coursera)
```


# Appendix
-----

### Variable Importance

```{r}
plot(varImp(fitMod))
```

### Selected Predictors vs Accuracy

```{r}
plot.train(fitMod)
```